### 2016-06-25 opsem.tex
RG section made succinct.
edits
Sec. 5 created.

\subsection{\txnimp: Syntax and Semantics}
\section{\txnimp}
\label{sec:opsem}

\label{sec:syntax}

@@

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations.  A session order
relation captures the sequential order of operations within a
transaction. In particular, it relates two effects, $\eta_1$ and
$\eta_2$, such that $\txn(\eta_1) = \txn(\eta_2)$ and $\id(\eta_1) <
\id(\eta_2)$.  The semantics assigns monotonically increasing
identifiers to effects, as defined by the $\id(\eta) > \maxId(\A)$
condition of \rulelabel{E-Aux}.  Evaluation contexts ($\ectx_i$) for
transaction-bound terms are defined so as to enforce a deterministic
sequential order of execution within a transaction, leading to a
deterministic total order among effect ids, which defines the session
order relation. Visibility ($\visZ$) on the other hand relates effects
across concurrent transactions.  Intuitively, $\visZ$ relates $\eta_1$
to $\eta_2$ if and only if $\eta_1$ was \emph{visible} to the
operation that generated $\eta_2$ during its execution, thus effecting
its return value ($\rval(\eta_2)$). For example, a read operation over
\C{X} may pick the value ($\rval$) of the write effect with highest id
among the visible effects (this is made possible by appropriately
defining $\interp{\cdot}$ in \rulelabel{E-Var}, as we show
later). Thus, the value of a read depends on what write effects it can
witness. An operation can only witness the effects of already
concluded operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition of
transactions. A more notable source of non-determinism, however, is
the \rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly consistent data stores. For instance, operations issued to an
eventually consistent (EC) replicated store could be dispatched to
different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores.

However, a machine that lets operations witness arbitrary subset of
the global state offers no isolation whatsoever. For example, it may
allow a read operation to witness writes of an uncommitted
transaction, violating RC isolation. Fortunately, our ability to
express an isolation semantics as constraints over happens-before
order through $\visZ$ and $\soZ$ relations, and the property of the
abstract machine to be parametric over the trace invariant ($\I$),
lets us solve this problem.  In particular, we continue to define
isolation semantics as constraints over $\visZ$ and $\soZ$, but
confine their domain of interpretation to the given trace so that they
now become trace well-formedness constraints. Well-formedness
constraints can be combined into a trace invariant ($\I$). Since the
abstract machine is guaranteed to preserve $\I$ at every step of the
execution, operations are guaranteed to experience the level of
isolation specified by $\I$.  Thus, in executions that run to
completion, the abstract machine models a store that provides the
required levels of isolation. Notably, the machine achieves this
without defining an operation semantics for isolation levels, solely
by relying on their declarative characterization as trace
well-formedness constraints.  \S\ref{sec:ansi-isolation} shows the
specification of various ANSI SQL isolation levels stated as trace
well-formedness constraints.

@@

\input{store-consistency}

>>>>

\section{\txnimp}
\label{sec:syntax}

\subsection{\txnimp: Syntax and Semantics}
\label{sec:opsem}

@@

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations, which shall
henceforth replace $\hbZ$ in the discussion.  A session order relation
captures the sequential order of operations within a transaction. In
particular, it relates two effects, $\eta_1$ and $\eta_2$, such that
$\txn(\eta_1) = \txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The
semantics assigns monotonically increasing identifiers to effects, as
defined by the $\id(\eta) > \maxId(\A)$ condition of
\rulelabel{E-Aux}.  Evaluation contexts ($\ectx_i$) for
transaction-bound terms are defined so as to enforce a deterministic
sequential order of execution within a transaction, leading to a
deterministic total order among effect ids, which defines the session
order relation. Visibility ($\visZ$) on the other hand relates effects
across concurrent transactions.  Intuitively, $\visZ$ relates $\eta_1$
to $\eta_2$ if and only if $\eta_1$ was \emph{visible} to the
operation that generated $\eta_2$ during its execution, thus effecting
its return value ($\rval(\eta_2)$). For example, a read operation over
\C{X} may pick the value ($\rval$) of the write effect with highest id
among the visible effects (this is made possible by appropriately
defining $\interp{\cdot}$ in \rulelabel{E-Var}, as we show later).
Thus, the value of a read depends on what write effects it can
witness. An operation can only witness the effects of already
concluded operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition of
transactions. A more notable source of non-determinism, however, is
the \rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly consistent data stores. For instance, operations issued to an
eventually consistent (EC) replicated store could be dispatched to
different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores. Stronger visiblity properties needed for stronger
isolation can however be recovered by imposing well-formedness
constraints over $\visZ$ via the trace invariant ($\I$). Since the
abstract machine is obligated to satisfy $\I$ at every step of the
execution, operations are guaranteed to experience the level of
isolation specified by $\I$.  Thus, in executions that run to
completion, the abstract machine models a store that provides the
required levels of isolation.  Notably, the machine achieves this
without defining an operation semantics for isolation levels, solely
by relying on their declarative characterization as trace
well-formedness constraints.  \S\ref{sec:ansi-isolation} shows the
specification of various ANSI SQL isolation levels stated as trace
well-formedness constraints.

% However, a machine that lets operations witness arbitrary subset of
% the global state offers no isolation whatsoever. For example, it may
% allow a read operation to witness writes of an uncommitted
% transaction, violating RC isolation. Fortunately, our ability to
% express an isolation semantics as constraints over happens-before
% order through $\visZ$ and $\soZ$ relations, and the property of the
% abstract machine to be parametric over the trace invariant ($\I$),
% lets us solve this problem.
% In particular, we continue to define isolation semantics as
% constraints over $\visZ$ and $\soZ$, but confine their domain of
% interpretation to the given trace so that they now become trace
% well-formedness constraints. Well-formedness constraints can be
% combined into a trace invariant ($\I$).

@@

### 2016-06-25 reasoning.tex
Sec. 5 created.

We now describe a proof system that lets us demonstrate that a \txnimp
program satisfies its high-level invariants in all small-step
executions that satisfy a chosen trace invariant\footnote{It is
important to mind the distinction between high-level invariants
(\eg~\C{X$\ge$0}) and trace invariants (\eg~\C{RC}). The former
constitute \emph{proof} \emph{obligations} for the programmer, whereas
the latter capture \emph{assumptions} about the machine. } ($\I$).
Since the trace invariant is a conjunction of store consistency
($\I_s$) and transaction isolation ($\I_c$) constraints, the
demonstration is a proof that the program preserves its invariants
when executed on a chosen store under a selection of isolation levels
for its transactions.

@@

Soundness of rely-guarantee judgments can now be stated
thus\footnote{Supplementary contains a rigorous proof of the soundness
theorem, and all relevant lemmas.}:
\begin{theorem}[Soundness] 
Rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to their semantics
defined above. 
\end{theorem}
\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
the rules of Fig.~\ref{fig:rg-rules}, then (a). every interleaved
multi-step reduction of $c$ starting from a trace that satisfies $P$
and $\I$, results in a trace that satisfies $Q$, and (b). the effect
that every small-step reduction of $c$ along the way has on the trace
is contained in $G$. Soundness of RG judgment for transaction-bound
commands ($\txnbox{c}_i$) is stated similarly.  For expressions,
soundness of the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also
proves that $e$ is always evaluated to a value that satisfies $C$.

\subsection{Soundness under Eventually Consistent Replication}

\subsection{Examples}

\GK{Here, I plan to include the full example that we dropped from
\S\ref{sec:motivation}.}

>>>>

We now describe a proof system that lets us demonstrate that a \txnimp
program satisfies its high-level invariants in all small-step
executions that satisfy a chosen trace invariant\footnote{It is
important to mind the distinction between high-level invariants
(\eg~\C{X$\ge$0}) and trace invariants (\eg~\C{RC}). The former
constitute \emph{proof} \emph{obligations} for the programmer, whereas
the latter capture \emph{assumptions} about the machine. } ($\I$).
Since the trace invariant captures transaction isolation constraints,
the demonstration is a proof that the given selection of isolation
levels for transactions is sufficient to enforce program's invariants.

@@

Soundness of rely-guarantee judgments can now be stated
thus\footnote{Supplementary contains a rigorous proof of the soundness
theorem, and all relevant lemmas.}:
\begin{theorem}[Soundness] 
Rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to their semantics
defined above. 
\end{theorem}
\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
the rules of Fig.~\ref{fig:rg-rules}, then (a). every interleaved
multi-step reduction of $c$ starting from a trace that satisfies $P$
and $\I$, results in a trace that satisfies $Q$, and (b). the effect
that every small-step of $c$ has on the trace is contained in $G$.
Soundness of RG judgment for transaction-bound commands
($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
is always evaluated to a value that satisfies $C$.
